\stktokens{}: Enforcing Well-bracketed Control Flow and Stack Encapsulation using Linear Capabilities

### 2018-10-18 fully-abstract-comp/popl2019/paper.tex
Move permission hierarchy figure
Various small improvements

%%% Capabilities
\begin{wrapfigure}{r}{0.26\linewidth}
  \centering
  \begin{tikzpicture}[main node/.style={}]
    \node[main node] (rwx) {$\rwx$};
    \node[main node] (rx) [below right of=rwx] {$\rx$};

@@

  \caption{Permission hierarchy}
  \label{fig:perm-hier}
\end{wrapfigure}
The concept of a capability is the cornerstone of any capability machine.
In its simplest form, a capability is a permission and a range of authority.
The permission dictates the operations the capability can be used for, and the range of authority specifies the range of memory it can act upon.
% Permissions and Range of authority
The capabilities on \trgcm{} are of the form $((\var{perm},\var{lin}),\var{base},\var{end},\var{addr})$ (defined in Figure~\ref{fig:target-syntax} with the rest of the syntax of \trgcm{}). Here $\var{perm}$ is the permission, and $[\var{base},\var{end}]$ is the range of authority.
The available permissions are read-write-execute ($\rwx$), read-write ($\rw$), read-execute ($\rx$), read-only ($\ro$), and null-permission ($\noperm$) ordered by $\le$ as illustrated in Figure~\ref{fig:perm-hier}.
% Pointer and linearity
In addition to the permission and range, capabilities also have a current address $\var{addr}$ and a linearity $\var{lin}$.
The linearity is either $\normal$ for traditional capabilities or $\linear$ for linear ones.
% What is linear capabilities, split/splice
A linear capability is a capability that cannot be duplicated.
This is enforced dynamically on the capability machine, so when a linear capability is moved between registers or memory, the source is cleared.
The non-duplicability of linear capabilities means that a linear capability cannot become aliased if it wasn't to begin with.

@@

              \inference{ \forall \Phi' \ldotp \neg (\Phi \step \Phi') \wedge \Phi' \neq \failed}

@@

% move w/ lin
The \texttt{move} instruction simply moves a word from one register to another.
It is, however, complicated slightly by the presence of the non-duplicable linear capabilities.
When a linear capability is moved, the source register must be cleared, so the capability is not duplicated.
This is done uniformly in the semantics using two approaches.
In the first one, we use the function $\linCons{}$ that returns $0$ for linear capabilities and is the identity for all other words.
When a word $w$ is transferred on the machine, then the source of $w$ is overwritten with $\linCons{w}$ which clears the source if $w$ was linear and leaves it unchanged otherwise.
In the case of \texttt{move}, the source register $r_2$ is overwritten with $\linCons{\Phi(r_2)}$.
We see the other approach later.

@@

% geta, cca
The instruction \texttt{geta} projects the current address (or seal) from a capability (or set of seals), and returns $-1$ for data and sealed capabilities.
\texttt{cca} changes the current address or seal of a capability or set of seals, respectively, by a given offset.
It handles linearity by updating the capability in-place, i.e.\ the source register is used as the target register.
This is the second approach we use to uniformly handle linearity.
The \texttt{jmp} instruction is a simple jump that just sets the $\pcreg$ register.

@@

The executable configuration describes the machine state, but it does not make it clear what components run on the machine and how they interact with each other.
To make this clearer, we introduce notions of components and programs from which we construct executable configurations.
A component is basically a program with entry points in the form of imports that need to be linked.
A component has exports that can satisfy the imports of other components.
The component definition is in Figure~\ref{fig:target-component-and-linking}.
% Structure
A base component $\comp_0$ consists of the following: a code memory segment, a data memory segment, a list of imported symbols, a list of exported symbols, two lists specifying the available seals\footnote{We will return to the seals in Section~\ref{sec:form-secur-with}.}, and a set of all the linear addresses (addresses governed by a linear capability).
The import list specifies where in memory imports should be placed, and imports are matched to exports via their symbols.
The exports are words each associated with a symbol.
%base = library
A base component has no main entry point, so it needs to be invoked by another component to run. For this reason, it can be seen as a library implementation.
A component is either a library component or a base component with a main in the form of a pair of sealed capabilities which can be seen as a program that still needs to be linked with libraries.
% Linking .
Components are combined into new components by linking them together.
Libraries can be linked together and incomplete programs can be linked with libraries, but two incomplete programs cannot be linked with each other.
Two components can be linked when their memories, seals, and linear addresses are disjoint.
They are combined by taking the union of each of their constituents.
For every import that is satisfied by an export of the other component, in the sense that they have the same symbol, the data memory is updated to have the exported word on the imported address.
The satisfied imports are removed from the import list in the resulting linked component and the exports are updated to be the exports of the two components.

@@

% Informally explain how we prevent this using linear capabilities
As the examples illustrate, the standard memory and object-like capabilities (here sealed capabilities) do not provide sufficient guarantees to enforce LSE and WBCF.
The problem is essentially that the stack and return pointers that a callee receives from a caller remain in effect after their intended lifetime: either when the callee has already returned or when they have themselves invoked other code. 
Linear capabilities offer a form of revocation\footnote{Revocation in the sense that if we hand out a linear capability and later get it back, then the receiver no longer has it or a copy of it as it is non-duplicable.} that can be used to prevent this from happening.

@@

% Prevention of attack 2
In a traditional calling convention with a single stack, the stack serves as a call stack keeping track of the order calls where made in and thus in which order they should be returned to.
A caller pushes a stack frame to the stack on call and a callee pops a stack frame from the stack upon return.
However without any enforcement, there is nothing to prevent a callee from returning on an arbitrary call on the call stack.
This is exactly what the adversary does in Figure~\ref{fig:ret-ptr-abuse} when they skip two stack frames.
In the presence of adversarial code, we need some enforcement mechanism that allows us to make sure that the order of the call stack is kept.
One way to enforce this would be to hand out a token on call that can only be used when the caller's stack frame is on top of the call stack.
The callee would have to provide this token on return to prove that it is allowed to return to the caller, and on return the token would be taken back by the caller to prevent it from being spent multiple times.
As it turns out, the stack capability for the unused part of the stack can be used as such a token in the following way:
On return the callee has to give back the stack capability they were given on invocation.
When the caller receives a stack capability back on return, they need to check that this token is actually spendable, i.e.\ check whether their stack frame is on top of the call stack.
They do this by attempting to restore the stack capability from before the call by splicing the return token with the stack capability for the local stack frame which at this point has been unsealed again.
If the splice is successful, then the caller knows that the two capabilities are adjacent. On the other hand, if the splice fails, then they are alerted to the fact that their stack frame may not be the topmost.
\stktokens{} uses this approach, and as illustrated in Figure~\ref{fig:ret-ptr-abuse-prev} it prevents the issue in Figure~\ref{fig:ret-ptr-abuse} as the adversary does not return a spendable token when they return.
\begin{figure}
  \centering
  \begin{subfigure}{0.4\linewidth}
    \centering
    \begin{tikzpicture}[scale=.5, every node={scale=.5}]
      % recurrent parts
      \draw (-0.5,13) node {\footnotesize 0};
      \draw (-0.5,12.5) edge[thin,-latex] (-0.5,11);
      \stdstackstart[13]
      \inactsf{(0,2)}{(4.5,4)} {\footnotesize Trusted\\ \footnotesize stack frame 1}
      \inactadv{(0,4)}{(4.5,7)} {\footnotesize Adv. stack frame 1}
      \inactsf{(0,7)}{(4.5,9)} {\footnotesize Trusted \\ \footnotesize stack frame 2}
      \actadv{(0,9)}{(4.5,12)} {\footnotesize Adv. stack frame 2}

@@

% Summary of CC
To summarise, our proposed calling convention \stktokens{} consists of the following measures:
\begin{enumerate}
\item \label{item:check-stkb} Check the base address of the stack capability before and after calls
\item \label{item:non-empty-sf} Make sure that local stack frames are non-empty
\item \label{item:return-data} Create token and data return capability on call: split the stack capability in two to get a stack capability for your local stack frame and a stack capability for the unused part of the stack. The former is sealed and used for the data part of the return pair.
\item \label{item:return-code} Create code return capability on call: Seal the old program pointer.
\item Reasonable use of seals: Return seals are only used to seal old program pointers, all return seals are only used for one call site, and they are not leaked.
\end{enumerate}
Item~\ref{item:check-stkb}-\ref{item:return-code} are captured by the code in Figure~\ref{fig:call-code} with the exception of checking stack base before calls. We do not include this check because it only needs to happen once between two calls which means that the check after a call suffices if the base of the stack pointer is not changed subsequently.

>>>>

%%% Capabilities
The concept of a capability is the cornerstone of any capability machine.
In its simplest form, a capability is a permission and a range of authority.
The permission dictates the operations the capability can be used for, and the range of authority specifies the range of memory it can act upon.
% Permissions and Range of authority
The capabilities on \trgcm{} are of the form $((\var{perm},\var{lin}),\var{base},\var{end},\var{addr})$ (defined in Figure~\ref{fig:target-syntax} with the rest of the syntax of \trgcm{}). Here $\var{perm}$ is the permission, and $[\var{base},\var{end}]$ is the range of authority.
The available permissions are read-write-execute ($\rwx$), read-write ($\rw$), read-execute ($\rx$), read-only ($\ro$), and null-permission ($\noperm$) ordered by $\le$ as illustrated in Figure~\ref{fig:perm-hier}.
% Pointer and linearity
In addition to the permission and range, capabilities also have a current address $\var{addr}$ and a linearity $\var{lin}$.
The linearity is either $\normal$ for traditional capabilities or $\linear$ for linear ones.
% What is linear capabilities, split/splice
A linear capability is a capability that cannot be duplicated.
This is enforced dynamically on the capability machine, so when a linear capability is moved between registers or memory, the source is cleared.
The non-duplicability of linear capabilities means that a linear capability cannot become aliased if it wasn't to begin with.
\begin{wrapfigure}{r}{0.26\linewidth}
  \centering
  \begin{tikzpicture}[main node/.style={}]
    \node[main node] (rwx) {$\rwx$};
    \node[main node] (rx) [below right of=rwx] {$\rx$};

@@

  \caption{Permission hierarchy}
  \label{fig:perm-hier}
\end{wrapfigure}

@@

              \inference{ \forall \Phi' \neq \failed \ldotp \Phi \nrightarrow \Phi'}

@@

% move w/ lin
The \texttt{move} instruction simply moves a word from one register to another.
It is, however, complicated slightly by the presence of the non-duplicable linear capabilities.
When a linear capability is moved, the source register must be cleared, so the capability is not duplicated.
This is done uniformly in the semantics using two approaches.
In the first one, we use the function $\linCons{}$ that returns $0$ for linear capabilities and is the identity for all other words.
When a word $w$ is transferred on the machine, then the source of $w$ is overwritten with $\linCons{w}$ which clears the source if $w$ was linear and leaves it unchanged otherwise.
In the case of \texttt{move}, the source register $\rn$ is overwritten with $\linCons{\Phi(\rn)}$.
We see the other approach later.

@@

% geta, cca
The instruction \texttt{geta} projects the current address (or seal) from a capability (or set of seals), and returns $-1$ for data and sealed capabilities.
\texttt{cca} (change current address) changes the current address or seal of a capability or set of seals, respectively, by a given offset.
It handles linearity by updating the capability in-place, i.e.\ the source register is used as the target register.
This is the second approach we use to uniformly handle linearity.
The \texttt{jmp} instruction is a simple jump that just sets the $\pcreg$ register.

@@

The executable configuration describes the machine state, but it does not make it clear what components run on the machine and how they interact with each other.
To clarify this, we introduce notions of components and programs from which we construct executable configurations.
A component is basically a program with entry points in the form of imports that need to be linked.
A component has exports that can satisfy the imports of other components.
The component definition is in Figure~\ref{fig:target-component-and-linking}.
% Structure
A base component $\comp_0$ consists of the following: a code memory segment, a data memory segment, a list of imported symbols, a list of exported symbols, two lists specifying the available seals\footnote{We will return to the seals in Section~\ref{sec:form-secur-with}.}, and a set of all the linear addresses (addresses governed by a linear capability).
The import list specifies where in memory imports should be placed, and imports are matched to exports via their symbols.
The exports are words each associated with a symbol.
%base = library
A base component has no main entry point, so it needs to be invoked by another component to run. For this reason, it can be seen as a library implementation.
A component is either a library component or a base component with a main in the form of a pair of sealed capabilities which can be seen as a program that still needs to be linked with libraries.
% Linking .
Components are combined into new components by linking them together.
Libraries can be linked together and incomplete programs can be linked with libraries, but two incomplete programs cannot be linked with each other.
Two components can be linked when their memories, seals, and linear addresses are disjoint.
They are combined by taking the union of each of their constituents.
For every import that is satisfied by an export of the other component, in the sense that they have the same symbol, the data memory is updated to have the exported word on the imported address.
The satisfied imports are removed from the import list in the resulting linked component and the exports are updated to be the exports of the two components.

@@

% Informally explain how we prevent this using linear capabilities
As the examples illustrate, the standard memory and object-like capabilities (here sealed capabilities) do not seem to provide sufficient guarantees to enforce LSE and WBCF.
The problem is essentially that the stack and return pointers that a callee receives from a caller remain in effect after their intended lifetime: either when the callee has already returned or when they have themselves invoked other code. 
Linear capabilities offer a form of revocation\footnote{Revocation in the sense that if we hand out a linear capability and later get it back, then the receiver no longer has it or a copy of it as it is non-duplicable.} that can be used to prevent this from happening.

@@

% Prevention of attack 2
In a traditional calling convention with a single stack, the stack serves as a call stack keeping track of the order calls where made in and thus in which order they should be returned to.
A caller pushes a stack frame to the stack on call and a callee pops a stack frame from the stack upon return.
However without any enforcement, there is nothing to prevent a callee from returning from an arbitrary call on the call stack.
This is exactly what the adversary does in Figure~\ref{fig:ret-ptr-abuse} when they skip two stack frames.
In the presence of adversarial code, we need some enforcement mechanism that allows us to make sure that the order of the call stack is kept.
One way to enforce this would be to hand out a token on call that can only be used when the caller's stack frame is on top of the call stack.
The callee would have to provide this token on return to prove that it is allowed to return to the caller, and on return the token would be taken back by the caller to prevent it from being spent multiple times.
As it turns out, the stack capability for the unused part of the stack can be used as such a token in the following way:
On return the callee has to give back the stack capability they were given on invocation.
When the caller receives a stack capability back on return, they need to check that this token is actually spendable, i.e.\ check whether their stack frame is on top of the call stack or not.
They do this by attempting to restore the stack capability from before the call by splicing the return token with the stack capability for the local stack frame which at this point has been unsealed again.
If the splice is successful, then the caller knows that the two capabilities are adjacent. On the other hand, if the splice fails, then they are alerted to the fact that their stack frame may not be the topmost.
\stktokens{} uses this approach; and as illustrated in Figure~\ref{fig:ret-ptr-abuse-prev}, it prevents the issue in Figure~\ref{fig:ret-ptr-abuse} as the adversary does not return a spendable token when they return.
\begin{figure}
  \centering
  \begin{subfigure}{0.4\linewidth}
    \centering
    \begin{tikzpicture}[scale=.5, every node={scale=.5}]
      % recurrent parts
      \draw (-0.5,13) node {\footnotesize 0};
      \draw (-0.5,12.5) edge[thin,-latex] (-0.5,11);
      \stdstackstart[13]
      \inactsf{(0,2)}{(4.5,4)} {\footnotesize Trusted\\ \footnotesize stack frame 1}
      \inactadv{(0,4)}{(4.5,7)} {\footnotesize Adv. stack frame 1}
      \inactsf{(0,7)}{(4.5,9)} {\footnotesize Trusted \\ \footnotesize stack frame 2}
      \actadv{(0,9)}{(4.5,12)} {\footnotesize Adv. stack frame 2}

@@

% Summary of CC
To summarise, our proposed calling convention \stktokens{} consists of the following measures:
\begin{enumerate}
\item \label{item:check-stkb} Check the base address of the stack capability before and after calls.
\item \label{item:non-empty-sf} Make sure that local stack frames are non-empty.
\item \label{item:return-data} Create token and data return capability on call: split the stack capability in two to get a stack capability for your local stack frame and a stack capability for the unused part of the stack. The former is sealed and used for the data part of the return pair.
\item \label{item:return-code} Create code return capability on call: Seal the old program pointer.
\item Reasonable use of seals: Return seals are only used to seal old program pointers, all return seals are only used for one call site, and they are not leaked.
\end{enumerate}
Item~\ref{item:check-stkb}-\ref{item:return-code} are captured by the code in Figure~\ref{fig:call-code} with the exception of checking stack base before calls. We do not include this check because it only needs to happen once between two calls which means that the check after a call suffices if the base of the stack pointer is not changed subsequently.
